Environmental pathogen reservoirs and habitat heterogeneity in a metapopulation

### 2014-04-23 Manuscript/manuscript.tex
More discussion edits
Added discussion of relationship to Hess, Gog, and Park papers and the
longevity dependence of the role of high and low quality patches.

Moreover, for a given longevity, these simulations suggest that high variance metapopulations are more prone to widespread pathogen spread than low variance metapopulations.  The mechanisms underlying this behavior can be understood by drawing parallels with models of pathogen spread through a contact network.  In this context, our metapopulation can be viewed a network with patches as nodes connected through immigration (with the strength of the connection measured by connectivity).  Many studies have shown that networks with high degree variance (where a node's degree is roughly equivalent to its total connectivity in our metapopulation model) are easily invaded by pathogens, which are then able to spread much more rapidly than in less heterogeneous networks (\cite{Pastor-Satorras2001}).

In addition to allowing us to explore the effects of longevity and variance independently, this experiment also allowed us to vary the two simultaneously and explore their interaction, suggesting that the specific effect of longevity on pathogen spread is mediated by the variance of the patch quality distribution.  In particular, metapopulations with high patch quality variance require lower environmental longevities to experience pandemic dynamics.  In this way, the effects of environmental longevity are exacerbated by a high variance patch quality distribution, especially for intermediate longevities.

This interaction between environmental longevity and the variance of the patch quality distribution is largely attributable to the presence of high quality patches.  Since a patch's connectivity is determined, at least in part, by its quality, high quality patches function as highly connected hubs in the metapopulation.  Specifically, high quality patches have both high in-degree (attract colonists from other patches) and a high out-degree (produce colonists that spread to other patches).  These two properties have important implications for pathogen spread in the metapopulation. The high in-degree helps position high quality patches as the ecological "traps" hypothesized above, wherein susceptible hosts repeatedly colonize high quality patches and subsequently become infected (Fig~\ref{infections}).  This infection pressure results in part from direct transmission from infected immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}), but is further augemented by environmental transmission.

This trap effect, coupled with the relatively low extinction rate on high quality patches helps to create a stable platform from which the pathogen can spread through the rest of the metapopulation.  Indeed, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}).  Once the pathogen infects high quality patches, it is able to spread to the rest of the metapopulation relatively easily.  This process helps to maintain infected patches throughout the metapopulation, which in turn feed back on high quality patches to ensure a steady stream of infected colonists that maintain the trap effect.  Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches serves to significantly increase pathogen spread.

Despite the dominating effects of high quality patches, low quality patches still play an important role in some cases by providing refuges on which susceptibles can escape infection. Even though individual low quality patches are unable to support consistent occupancy, at the metapopulation level, the presence of low quality patches increases overall susceptible occupancy, an effect that is strongest at intermediate longevities.   This phenomenon represents the other side of the hub role played by high quality patches in that low quality patches are relatively weakly connected to the metapopulation and are therefore infrequently colonized by infected individuals.  Moreover, the instability of low quality patches (i.e. the high extinction rate), together with the increased mortality of infected populations, means that infected populations do not persist long on low quality patches.  Because low quality patches are so infrequently colonized, the environmental reservoir left behind by these infected populations generally decays before it has the opportunity to infect new susceptible colonists.

Due to their lower extinction rates, high quality patches are able to maintain stable occupancy despite widespread infection.  On the other hand, at high pathogen longevities, the positive effect of low qualtiy distributions on susceptible occupancy is overwhelmed, and low quality patches are unable to support infected occupancy, resulting in an increase in host extinction events.  Thus, the role of the patch quality distribution and the presence of high or low quality patches is generally greatest for low to intermediate pathogen longevity -- once longevity is high enough, the pathogen persists and spreads easily enough that the effects of individual patches are washed out.  Then the role of patch quality lies primarily in its effect on extinction rates, and whether infected hosts can persist.

In metapopulations without disease, high quality patches generally serve as sources, while low quality patches are sinks.  The presence of a pathogen then considerably alters, and even inverts, these patterns.  Not only do high quality patches become sinks in the presence of a pathogen with an environmental reservoir, they become ecological traps, which can be even more detrimental to the overall population (\cite{Kristan2003}).  From a management perspective, it is critical to understand this shift, as previous theoretical work suggests that management should focus on patches where conditions are most favorable (\cite{Strasser2010}).  Generally, this means that maintaining high quality habitat should form the focus of managers, but Strasser et al. (2010) show that stochastic disturbance (e.g. disease) can lead to cases where focusing on low quality patches is more effective in increasing population growth rate.

Habitat fragmentation and disease represent two major threats to wildlife populations.  As we have seen, these two factors can interact in complex and interesting ways, especially when the pathogen is able to persist in the environment.  Failing to consider these interactions and their effect on the roles played by high and low quality habitat overlooks the importance of low-quality habitat for the persistence of wildlife populations and the importance of high-quality habitat for the persistence of wildlife disease.

>>>>

Moreover, for a given longevity, these simulations suggest that high variance metapopulations are more prone to widespread pathogen spread than low variance metapopulations (Fig~\ref{poutcome}.  The mechanisms underlying this behavior can be understood by drawing parallels with models of pathogen spread through a contact network.  In this context, our metapopulation can be viewed a network with patches as nodes connected through immigration (with the strength of the connection measured by connectivity).  Many studies have shown that networks with high degree variance (where a node's degree is roughly equivalent to its total connectivity in our metapopulation model) are easily invaded by pathogens, which are then able to spread much more rapidly than in less heterogeneous networks (\cite{Pastor-Satorras2001}).

These simulations also offer insight into the interaction between environmental longevity and habitat heterogeneity, suggesting that the specific effect of longevity on pathogen spread is mediated by the variance of the patch quality distribution.  In particular, metapopulations with high patch quality variance require lower environmental longevities to experience pandemic dynamics (Fig~\ref{poutcome}).  In this way, the effects of environmental longevity are exacerbated, or amplified, by a high variance patch quality distribution, especially at intermediate longevities.

This interaction between environmental longevity and the variance of the patch quality distribution is largely driven by the presence of high quality patches.  Since a patch's connectivity is determined, at least in part, by its quality, high quality patches function as highly connected hubs in the metapopulation.  Specifically, high quality patches both attract more colonists from other patches and produce more colonists that spread to other patches.  These two properties have important implications for pathogen spread in the metapopulation. Attracting colonists helps position high quality patches as the ecological "traps" hypothesized above, wherein susceptible hosts repeatedly colonize high quality patches and subsequently become infected (Fig~\ref{infections}).  This infection pressure results in part from direct transmission from infected immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}), but is further augemented by environmental transmission.

This trap effect, coupled with the relatively low extinction rate on high quality patches helps to create a stable platform from which the pathogen can spread through the rest of the metapopulation.  Indeed, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}).  Once the pathogen infects high quality patches, the larger number of colonists produced allow it to spread to the rest of the metapopulation relatively easily.  This process helps to maintain infected occupancy throughout the metapopulation, which in turn feed back on high quality patches to ensure a steady stream of infected colonists that maintain the trap effect.  Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches serves to significantly increase pathogen spread and infected occupancy in the metapopulation (Fig~\ref{sens}).

Despite the dominating effects of high quality patches, low quality patches still play an important role in some cases by facilitating increased susceptible occupancy (Fig~\ref{sens}). Even though individual low quality patches are unable to support consistent occupancy (Fig~\ref{simvis}), at the metapopulation level, the presence of low quality patches increases overall susceptible occupancy relative to a lower variance metapopulation, an effect that is strongest at intermediate longevities (Fig~\ref{sens}).   This phenomenon represents the other side of the hub role played by high quality patches in that low quality patches are relatively weakly connected to the metapopulation and are therefore infrequently colonized by infected individuals.  Moreover, the instability of low quality patches (i.e. the high extinction rate), together with the increased mortality of infected populations, means that infected populations do not persist long on low quality patches.  Because low quality patches are so infrequently colonized, at low to intermediate longevities, the environmental reservoir left behind by these infected populations generally decays before it has the opportunity to infect new susceptible colonists.  As a result, low quality patches represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and thereby reducing infected occupancy (Fig~\ref{sens}).

However, the importance of the patch quality distribution in pathogen spread depends on the pathogen's longevity, with the strongest effects of patch quality observed for intermediate longevities.  For these longevities, the timescale of pathogen decay roughly matches the timescale of the colonization and extinction processes.  In contrast, for low longevities, when the pathogen reservoir decays much faster than colonizations and extinctions occur, environmental transmission has little effect on disease dynamics and the effect of patch quality is relatively weak.  Conversely, when the pathogen decays much slower than colonizations and extinctions occur, i.e. high longevity, the pathogen reservoir persists on nearly all patches and the effects of patch quality are washed out.  In these cases, the role of patch quality lies primarily in its effect on extinction rates.  Due to their lower extinction rates, high quality patches are able to maintain stable occupancy despite widespread (pandemic) infection, while low quality patches are unable to support infected occupancy, resulting in an increase in host extinction events at the highest longevities.

These findings suggest the presence of a trade-off between facilitating metapopulation stability and inhibiting disease spread, with a pathogen's environmental longevity controlling the roles of different quality patches with respect to this trade-off.  At low to intermediate longevities, the addition of high quality patches functions similarly to increasing movement in the model developed by \cite{Hess1996}, increasing pathogen spread and reducing overall occupancy (Fig~\ref{sens}c).  In these cases, rather than serving as sources in the metapopulation, high quality patches  become sinks, and more importantly, they become ecological traps, which can be even more detrimental to the overall population (\cite{Kristan2003}).  From a management perspective, it is critical to understand this shift, as previous theoretical work suggests that management should focus on patches where conditions are most favorable (\cite{Strasser2010}).  Generally, this means that maintaining high quality habitat should form the focus of managers, but Strasser et al. (2010) show that stochastic disturbance (e.g. disease) can lead to cases where focusing on low quality patches is more effective in increasing population growth rate.

However, as pathogen longevity increases, the trade-off shifts and the addition of high quality patches is a net benefit to overall occupancy, reflecting the findings of \cite{Gog2002}, who find that under background infection from an alternative host, increasing movement rarely reduces occupancy, despite facilitating pathogen spread.  \cite{Park2012} further develops this work by noting that the magnitude of the detrimental effects of increased moevement on occupancy depends on the relative strength of different transmission pathways.  Specifically, he finds that as environmental transmission becomes stronger (e.g. higher longevity) relative to direct transmission, then total occupancy increases monotonically with movement, i.e. movement is a net benefit to the metapopulation.  These studies, coupled with the results presented here, suggest that when there is a persistent source of infection throughout the metapopulation (e.g. when the pathogen is long-lived in the environment), factors that facilitate metapopulation stability (i.e. increased movement, high quality patches) are a greater benefit than factors that inhibit pathogen spread (i.e. decreased movement, low quality patches).

Spatial structure and disease spread represent two major factors influencing the dynamics of wildlife populations.  As we have seen, in cases where the pathogen is able to persist and remain infectious in the environment, these two factors can interact in complex and interesting ways.  Failing to consider these interactions and their effect on the roles played by high and low quality habitat risks overlooking the importance of low-quality habitat for the persistence of wildlife populations and the importance of high-quality habitat for the persistence of wildlife disease.
